Adaptive Safe Strong Screening for Efficient Lasso-Type Problems Optimization over the Solution Path

### 2021-06-23 main.tex
Address reviewer issues except 1

Throughout this paper, it will be assumed that $y$ is centered and $\{x_j\}_{j=1}^p$ are centered and standardized:
\begin{equation}
  \label{eq:std}
  \sum_{i=1}^ny_i=0, \qquad \sum_{i=1}^n x_{ij}=0, \qquad \sum_{i=1}^n x_{ij}^2=n,\qquad j=1,2,...,p.
\end{equation}
Standardization ensures that the penalty is applied equally to all features, even if originally measured in different units. It also improves the efficiency and stability of the optimization, and helps to simplify optimization.  For example, under this standardization, the intercept is exactly 0 and thus may be ignored.  Standardization also simplifies the formulas for screening rules and thus allows for more efficient implementation.  Furthermore, this can be done without loss of generality as one can always obtain the solution on the original scale through a simple linear transformation of the standardized solution.

@@

The EDPP rules are derived from the dual formulation of the lasso problem and provide a bound for the residual vector $r(\lambda)$. Then using the KKT conditions, some features can be guaranteed to have zero coefficients and be discarded safely. Two forms of the EDPP rule, basic EDPP (BEDPP) and sequential EDPP (SEDPP)\citep{wang2013lasso}, were proposed and under the standardization \eqref{eq:std} can be simplified in the following theorems \citep{Zeng2021}:

@@

Compared to one-step methods like BEDPP, the reference solutions are continually updated.  As a result, the reference is always relatively close to the solution, and screening remains powerful over the entire path, unlike one-step methods where screening power drops drastically as $\lambda_l$ gets further away from $\lambda_0$. Compared to sequential methods like SEDPP, the adaptive method only requires $O(np)$ computations per update, while SEDPP requires $O(np)$ computations for each $\lambda$ value. The framework of adaptive safe screening can also be easily applied to other sequential safe rules, to form a family of adaptive safe rules.

@@

            \uIf{Post-convergence check does not pass}{
                Add violating features to $\mathcal{R}_{l}$\;
            }
        }
        \uIf{Update conditions met}{
            Reference = $\lambda_{l}$

@@

In this section we will fill in line 11 in Algorithm~\ref{alg:path}. We develop an algorithm to determine when to update the reference for screening. This decision is guided by evaluating screening power at previous $\lambda$ values, and is therefore adaptive, updating the reference more or less frequently as we continue along the solution path. Our goal is to minimize the computation cost, so we will first analyse computation complexity for each part in Algorithm~\ref{alg:path}. Then we will propose an adaptive algorithm that can minimize the entire computation cost.

@@

This complexity analysis will only consider calculations with $O(np)$ complexity. The remaining calculations, those with $O(p)$ complexity for example, are not only of lower order, but also consume less memory and thus usually do no require out-of-core computing. Calculation of $\{x_j^Ty\}_{j=1}^p$ is needed only once for the whole solution path; we ignore this as well as its computational cost is the same across all approaches. We assume that strong rule screening is always powerful in the sense that its set of non-discarded features, $\mathcal{R}_{l+b}$, satisfies $|\mathcal{R}_{l+b}|=o(p) \approx 0$, but that safe rule screening, with corresponding set $\mathcal{S}_{l+b}$, may not be. We also assume that post-convergence check will usually pass in the first iteration and thus is only needed approximately once per $\lambda$.

Under these assumptions, there are two dominating parts in the algorithm within each update. First, the computation of $\{x_j^Tr(\lambda_l)\}_{j=1}^p$ at the time of updating the reference, which requires $np$ multiplications. Then, for each $\lambda_{l+b}$, the post-convergence check requires approximately $|\mathcal{S}_{l+b} \setminus \mathcal{R}_{l+b}| \approx |\mathcal{S}_{l+b}|$ KKT condition checks, each of which requires $n$ multiplications. These sum to $n\sum_{b=1}^{B}|\mathcal{S}_{l=b}|$ multiplications, where $l+B$ is the start of the next update. Combining these two parts, the average cost per $\lambda$ in the block $\lambda_{l+1},...,\lambda_{l+B}$ is
\begin{equation}
    \label{eq:cost}
    \bar{T}_l(B) \approx n\frac{p}{B}+n\frac{\sum_{b=1}^B|\mathcal{S}_{l+b}|}{B},
\end{equation}
where the first term represents the cost of updating the safe screening rule and the second term the cost of KKT checking. Delaying the next update (increasing $B$) means reusing calculations additional times, thus reducing the cost of safe screening (first term). However, reusing an old reference reduces the power of the safe rule, thereby increasing the cost of KKT checking (second term).

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.6\textwidth]{plots/batchsizes.eps}    \caption{Proportion of features rejected by screening rules vs index of $\lambda$-values for different update delay.}
    \label{fig:batchsizes}
\end{figure}

BEDPP is the special case of adaptive EDPP that never updates ($B=L$). When the reference is not updated, safe rule screening will eventually fail to discard any features and the KKT condition checking cost will increase to its maximum, $np$. SEDPP is the special case that updates at every $\lambda$ value ($B=1$). It remains very powerful along the path, making the KKT condition checking cost small, but safe screening updates themselves are costly and require $np$ operations at every $\lambda$-value. Adaptive EDPP can strike a balance between BEDPP and SEDPP. In this example, simply using a fixed batch size $B=10$ can reduce the safe rule updating cost by 90\% compared to SEDPP and reduce the KKT checking cost by about 80\% compared to BEDPP.

@@

We use this updating condition at line 11 in Algorithm~\ref{alg:path} to form a complete pathwise lasso algorithm with adaptive hybrid screening.

@@

Consider the sparse logistic regression model where estimates of the coefficient vector $\beta$ and intercept $\beta_0$ are obtained by minimizing
\begin{equation}
    \label{eq:logis}
    %\hat{\beta}(\lambda)=\underset{(\beta,\beta_0)\in \mathbb{R}^p\times\mathbb{R}^1}{\mathrm{argmin}}-
  \frac{1}{n} \sum_{i=1}^n \big\{y_i\log\pi_i(\beta) + (1-y_i)\log(1 -\pi_i(\beta))\big\} + \lambda||\beta||_1,
\end{equation}
where $\pi(\beta)=\{\pi_i(\beta)\}_{i=1}^n=\{p(y_i=1|x_{i1},x_{i2},...,x_{ip})\}_{i=1}^n=\{e^{\eta_i(\beta)}/(1+e^{\eta_i(\beta)})\}_{i=1}^n$ is the predicted value vector and $\eta(\beta)=X\beta+\beta_0\mathbf{1}$ is the linear function. The quantity $X$ is still the $n\times p$ standardized feature matrix but in this model $y\in\{0,1\}^n$ is an unstandardized binary response vector; unlike linear regression, the intercept cannot be eliminated through centering in logistic regression.

@@

When computing the adaptive Slores, $\{x_j^Ty\}_{j=1}^p$ needs to be computed only once, after which it can be stored and reused for the whole solution path. The quantity $\{x_j^Tx_*\}_{j=1}^p$ can be similarly reused unless $x_*$ changes, which happens if the corresponding regression coefficient drops out of the model; this is relatively rare, but can happen. Other than those two terms, the only term that requires expensive $O(np)$ computations is $\{x_j^Tr(\lambda_l)\}_{j=1}^p$, which needs to be computed only when we update the reference solution. Thus Algorithm~\ref{alg:path} can be applies to the sparse logistic model, with Slores replacing EDPP, to form the adaptive hybrid rule algorithm for sparse logistic regression.

@@

\begin{theorem}[Adaptive EDPP for group lasso]
  For the group lasso problem \eqref{eq:glasso}, suppose $\lambda_0>\lambda_1>...>\lambda_L$, with $\lambda_0:= \max_j ||X_g^Ty||_2 / (n\sqrt{p_g})$. Under conditions \eqref{eq:std} and \eqref{eq:stdg},
    \begin{enumerate}
        \item For any $l=1,2,...,L-1$, given $\hat{\beta}(\lambda_l)$, $r(\lambda_l)$ and $\hat{y}(\lambda_l):=\sum_{g=1}^GX_g\hat{\beta}_g(\lambda_l)$, then for some next update point $l+B$ and any $b$ s.t. $l<l+b\leq l+B\leq L$, $\hat{\beta}_g(\lambda_{l+b})=0$ if
        \begin{equation}
            \begin{split}
                &\left|\left|2\lambda_{l+b}X_g^Tr(\lambda_l)+(\lambda_l-\lambda_{l+b})\left( X_g^Ty-\frac{y^T\hat{y}(\lambda_l)X_g^T\hat{y}(\lambda_l)}{||\hat{y}(\lambda_l)||_2^2}\right)\right|\right|_2\\
                &<2n\lambda_l\lambda_{l+b}\sqrt{p_g}-(\lambda_l-\lambda_{l+b})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_l))^2}{||\hat{y}(\lambda_l)||_2^2}}.
            \end{split}
        \end{equation}
        \item For l=0, i.e., $\lambda_l=\lambda_0$, let $X_*=\arg \max_{X_g} ||X_g^Ty||_2 / (n\sqrt{p_g})$ with dimension $n\times p_*$, then for some next update point $B$ and any $b$ s.t. $0<b\leq B\leq L$, $\hat{\beta}_g(\lambda_{b})=0$ if
        \begin{equation}
        \left|\left|(\lambda_0+\lambda_b)X_g^Ty-(\lambda_0-\lambda_b)X_g^TX_*X_*^Ty\right|\right|_2<2n\lambda_0\lambda_b\sqrt{p_g}-(\lambda_0-\lambda_b)\sqrt{n||y||_2^2-n^2\lambda_0^2p_*}.
    \end{equation}
    \end{enumerate}
\end{theorem}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{plots/511.eps}    \caption{Comparing speed of screening methods for lasso model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to AC. First column: varying number of features $p$. Second column: varying sample size $n$. Third column: varying signal strength $\beta_{\max}$. Fourth column: varying sparsity $s$.}
    \label{fig:5.1.1a}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.82\textwidth]{plots/511b.eps}    \caption{Comparing speed of screening methods for lasso model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to AC. First column: Varying number of grid points $L$. Second column: Varying range of grid $\lambda_{\min}/\lambda_{\max}$. Third column: Varying number of grid points $L$ with fixed spacing.}
    \label{fig:5.1.1b}
\end{figure}

Figure~\ref{fig:5.1.1a} and~\ref{fig:5.1.1b} shows the computation time (top row) and relative computation time compared to AC (bottom row) of solving lasso problem with different screening methods. AHR is uniformly the fastest under all settings, typically 2-8 times faster than any other screening method. The difference is largest when the data set is large (large $n$ or large $p$), the active features are highly sparse (small $s$) or the grid of $\lambda$-value is fine (large $L$ or $\lambda_{\min}/\lambda_{\max}$). For example, when $L=1000$, AHR is 18 times faster than any other approach.

@@

The proposed adaptive hybrid screening framework is flexible and can be extended in several ways. The framework is general and does not require a specific algorithm, strong rule, or safe rule. As faster algorithms to solve the lasso problem, better safe rules, or improved strong rules are developed, these can be directly plugged into the adaptive hybrid framework with a corresponding improvement in computational efficiency. Nevertheless, there are still several lasso type problems without sequential safe rules, such as sparse Cox regression, sparse Poisson regression, and sparse support vector machines. As safe rules for these problems are developed, they can be readily fit into our framework to produce efficient algorithms as well.

>>>>

Throughout this paper, it will be assumed for simplicity that $y$ is centered and $\{x_j\}_{j=1}^p$ are centered and standardized:
\begin{equation}
  \label{eq:std}
  \sum_{i=1}^ny_i=0, \qquad \sum_{i=1}^n x_{ij}=0, \qquad \sum_{i=1}^n x_{ij}^2=n,\qquad j=1,2,...,p.
\end{equation}
Standardization ensures that the penalty is applied equally to all features, even if originally measured in different units. It also improves the efficiency and stability of the optimization, and helps to simplify optimization. For example, under this standardization, the intercept is exactly 0 and thus may be ignored. Standardization also simplifies the formulas for screening rules and thus allows for more efficient implementation. Furthermore, this can be done without loss of generality as one can always obtain the solution on the original scale through a simple linear transformation of the standardized solution. Nevertheless, the methods proposed in this paper also work without standardization.

@@

The EDPP rules are derived from the dual formulation of the lasso problem and provide a bound for the residual vector $r(\lambda)$. Then using the KKT conditions, some features can be guaranteed to have zero coefficients and be discarded safely. Two forms of the EDPP rule, basic EDPP (BEDPP) and sequential EDPP (SEDPP)\citep{wang2013lasso}, were proposed without standardization and under the standardization \eqref{eq:std} can be simplified in the following theorems \citep{Zeng2021}:

@@

Compared to one-step methods like BEDPP, the reference solutions are updated systematically. As a result, the reference is always relatively close to the solution, and screening remains powerful over the entire path, unlike one-step methods where screening power drops drastically as $\lambda_l$ gets further away from $\lambda_0$. Compared to sequential methods like SEDPP, the adaptive method only requires $O(np)$ computations per update, while SEDPP requires $O(np)$ computations for each $\lambda$ value. The framework of adaptive safe screening can also be easily applied to other sequential safe rules, to form a family of adaptive safe rules.

@@

            \uIf{Post-convergence check does not pass}{
                Add violating features to $\mathcal{R}_{l}$\;
            }
        }
        \uIf{Update conditions met\label{line:upcond}}{
            Reference = $\lambda_{l}$

@@

In this section we will fill in line \ref{line:upcond} in Algorithm~\ref{alg:path}. We develop an algorithm to determine when to update the reference for screening. This decision is guided by evaluating screening power at previous $\lambda$ values, and is therefore adaptive, updating the reference more or less frequently as we continue along the solution path. Our goal is to minimize the computation cost, so we will first analyse computation complexity for each part in Algorithm~\ref{alg:path}. Then we will propose an adaptive algorithm that can minimize the entire computation cost.

@@

This complexity analysis will only consider calculations with $O(np)$ complexity. The remaining calculations, those with $O(p)$ complexity for example, are not only of lower order, but also consume less memory and thus usually do no require out-of-core computing. Calculation of $\{x_j^Ty\}_{j=1}^p$ is needed only once for the whole solution path; we ignore this as well as its computational cost is the same across all approaches. We assume that strong rule screening is always powerful in the sense that its set of non-discarded features, $\mathcal{R}_{l+b}$, satisfies $|\mathcal{R}_{l+b}|=o(p) \approx 0$, but that safe rule screening, with corresponding set $\mathcal{S}_{l+b}$, may not be. We also assume that post-convergence check will usually pass in the first iteration and thus is only needed approximately once per $\lambda$. These two assumptions are supported by experiments from both the origin paper of SSR \citep{Tibshirani2012} and many other papers studying feature screening.

Under these assumptions, there are two dominating parts in the algorithm within each update. First, the computation of $\{x_j^Tr(\lambda_l)\}_{j=1}^p$ at the time of updating the reference, which requires $np$ multiplications. Then, for each $\lambda_{l+b}$, the post-convergence check requires approximately $|\mathcal{S}_{l+b} \setminus \mathcal{R}_{l+b}| \approx |\mathcal{S}_{l+b}|$ KKT condition checks, each of which requires $n$ multiplications. These sum to $n\sum_{b=1}^{B}|\mathcal{S}_{l+b}|$ multiplications, where $l+B$ is the start of the next update. Combining these two parts, the average cost per $\lambda$ in the block $\lambda_{l+1},...,\lambda_{l+B}$ is
\begin{equation}
    \label{eq:cost}
    \bar{T}_l(B) \approx n\frac{p}{B}+n\frac{\sum_{b=1}^B|\mathcal{S}_{l+b}|}{B},
\end{equation}
where the first term represents the cost of updating the safe screening rule and the second term the cost of KKT checking. Delaying the next update (increasing $B$) means reusing calculations additional times, thus reducing the cost of safe screening (first term). However, reusing an old reference reduces the power of the safe rule, thereby increasing the cost of KKT checking (second term).

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.6\textwidth]{batchsizes.eps}    \caption{Proportion of features rejected by screening rules vs index of $\lambda$-values for different update delay.}
    \label{fig:batchsizes}
\end{figure}

BEDPP is the special case of adaptive EDPP that never updates ($B=L$). When the reference is not updated, safe rule screening will eventually fail to discard any features and the KKT condition checking cost will increase to its maximum, $np$. SEDPP is the special case that updates at every $\lambda$ value ($B=1$). It remains very powerful along the path, making the KKT condition checking cost small, but safe screening updates themselves are costly and require $np$ operations at every $\lambda$-value. Adaptive EDPP can strike a balance between BEDPP and SEDPP. The screening power drops as the same reference keeps being reused, indicated by the decreasing part of the green line. However, the reference gets updated periodically and at those points screening regains strength, indicated by the upward jumps on the green line. The overall screening power is kept relative high with only a few updates. In this example, simply using a fixed batch size $B=10$ can reduce the safe rule updating cost by 90\% compared to SEDPP and reduce the KKT checking cost by about 80\% compared to BEDPP.

@@

We use this updating condition at line \ref{line:upcond} in Algorithm~\ref{alg:path} to form a complete pathwise lasso algorithm with adaptive hybrid screening.

@@

Consider the sparse logistic regression model where estimates of the coefficient vector $\beta$ and intercept $\beta_0$ are obtained by minimizing
\begin{equation}
    \label{eq:logis}
    %\hat{\beta}(\lambda)=\underset{(\beta,\beta_0)\in \mathbb{R}^p\times\mathbb{R}^1}{\mathrm{argmin}}
  -\frac{1}{n} \sum_{i=1}^n \big\{y_i\log\pi_i(\beta) + (1-y_i)\log(1 -\pi_i(\beta))\big\} + \lambda||\beta||_1,
\end{equation}
where $\pi(\beta)=\{\pi_i(\beta)\}_{i=1}^n=\{p(y_i=1|x_{i1},x_{i2},...,x_{ip})\}_{i=1}^n=\{e^{\eta_i(\beta)}/(1+e^{\eta_i(\beta)})\}_{i=1}^n$ is the predicted value vector and $\eta(\beta)=X\beta+\beta_0\mathbf{1}$ is the linear function. The quantity $X$ is still the $n\times p$ standardized feature matrix but in this model $y\in\{0,1\}^n$ is an unstandardized binary response vector; unlike linear regression, the intercept cannot be eliminated through centering in logistic regression.

@@

When computing the adaptive Slores, $\{x_j^Ty\}_{j=1}^p$ needs to be computed only once, after which it can be stored and reused for the whole solution path. The vector $x_*$ can be chosen arbitrarily from features that has non-zero coefficients at the reference $\lambda_l$, for example, the feature whose coefficient that first enters the model. The quantity $\{x_j^Tx_*\}_{j=1}^p$ can be similarly reused unless the corresponding regression coefficient drops out of the model; this can happen but is relatively rare. Other than those two terms, the only term that requires expensive $O(np)$ computations is $\{x_j^Tr(\lambda_l)\}_{j=1}^p$, which needs to be computed only when we update the reference solution. Thus Algorithm~\ref{alg:path} can be applies to the sparse logistic model, with Slores replacing EDPP, to form the adaptive hybrid rule algorithm for sparse logistic regression.

@@

\begin{theorem}[Adaptive EDPP for group lasso]
  For the group lasso problem \eqref{eq:glasso}, suppose $\lambda_0>\lambda_1>...>\lambda_L$, with $\lambda_0:= \max_g ||X_g^Ty||_2 / (n\sqrt{p_g})$. Under conditions \eqref{eq:std} and \eqref{eq:stdg},
    \begin{enumerate}
        \item For any $l=1,2,...,L-1$, given $\hat{\beta}(\lambda_l)$, $r(\lambda_l)$ and $\hat{y}(\lambda_l):=\sum_{g=1}^GX_g\hat{\beta}_g(\lambda_l)$, then for some next update point $l+B$ and any $b$ s.t. $l<l+b\leq l+B\leq L$, $\hat{\beta}_g(\lambda_{l+b})=0$ if
        \begin{equation}
            \begin{split}
                &\left|\left|2\lambda_{l+b}X_g^Tr(\lambda_l)+(\lambda_l-\lambda_{l+b})\left( X_g^Ty-\frac{y^T\hat{y}(\lambda_l)X_g^T\hat{y}(\lambda_l)}{||\hat{y}(\lambda_l)||_2^2}\right)\right|\right|_2\\
                &<2n\lambda_l\lambda_{l+b}\sqrt{p_g}-(\lambda_l-\lambda_{l+b})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_l))^2}{||\hat{y}(\lambda_l)||_2^2}}.
            \end{split}
        \end{equation}
        \item For l=0, i.e., $\lambda_l=\lambda_0$, let $X_*=\arg \max_{X_g} ||X_g^Ty||_2 / (n\sqrt{p_g})$ with dimension $n\times p_*$, then for some next update point $B$ and any $b$ s.t. $0<b\leq B\leq L$, $\hat{\beta}_g(\lambda_{b})=0$ if
        \begin{equation}
        \left|\left|(\lambda_0+\lambda_b)X_g^Ty-(\lambda_0-\lambda_b)X_g^TX_*X_*^Ty\right|\right|_2<2n\lambda_0\lambda_b\sqrt{p_g}-(\lambda_0-\lambda_b)\sqrt{n||y||_2^2-n^2\lambda_0^2p_*}.
    \end{equation}
    \end{enumerate}
\end{theorem}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{511.eps}    \caption{Comparing speed of screening methods for lasso model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to AC. First column: varying number of features $p$. Second column: varying sample size $n$. Third column: varying signal strength $\beta_{\max}$. Fourth column: varying sparsity $s$.}
    \label{fig:5.1.1a}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.82\textwidth]{511b.eps}    \caption{Comparing speed of screening methods for lasso model under different settings. Top row: computation time in second. Bottom row: relative computation time compared to AC. First column: Varying number of grid points $L$. Second column: Varying range of grid $\lambda_{\min}/\lambda_{\max}$. Third column: Varying number of grid points $L$ with fixed spacing.}
    \label{fig:5.1.1b}
\end{figure}

Figure~\ref{fig:5.1.1a} and~\ref{fig:5.1.1b} shows the computation time (top row) and relative computation time compared to AC (bottom row) of solving lasso problem with different screening methods. AHR is uniformly the fastest under all settings, typically 2-8 times faster than any other screening method. The difference is largest when the data set is large (large $n$ or large $p$), the active features are highly sparse (small $s$) or the grid of $\lambda$-value is fine (large $L$ or $\lambda_{\min}/\lambda_{\max}$). For example, when $L=1000$, AHR is 18 times faster than any other approach. When $\lambda_{\min}/\lambda_{\max}$ is large, both AHR and HSSR can take advantage because all the $\lam$ values in the path are close to the first reference $\lambda_{\max}$.

@@

The proposed adaptive hybrid screening framework is flexible and can be extended in several ways. The framework is general and does not require a specific algorithm, strong rule, or safe rule. As faster algorithms to solve the lasso problem, better safe rules, or improved strong rules are developed, these can be directly plugged into the adaptive hybrid framework with a corresponding improvement in computational efficiency. As a result of the flexibility, the standardization conditions required in this paper can also be removed if we substitute the underlying sequential safe rules with the forms in their original papers that do not require standardization. Nevertheless, there are still several lasso type problems without sequential safe rules, such as sparse Cox regression, sparse Poisson regression, and sparse support vector machines. The sequential strong rule can address these problems, but as safe rules for these problems are developed, our framework can readily fit both strong rules and safe rules in to produce an even more efficient algorithms.
